---
title: "Homework 2"
output:
  blogdown::html_page:
    highlight: tango
---

```{r echo = FALSE}
knitr::opts_chunk$set(eval = FALSE)
```

## Goals

1. Import counts from mouse E18 brain scATACseq (5K) (provided by 10X Genomics) in R and process them  

> Processed data for mouse E18 brain scATACseq can be obtained directly from [10X Genomics](https://support.10xgenomics.com/single-cell-atac/datasets/1.2.0/atac_v1_E18_brain_fresh_5k)

2. Transfer annotations from mouse E18 brain scRNAseq (5K) to mouse E18 brain scATACseq

> Processed data for mouse E18 brain scRNAseq can be obtained directly from [10X Genomics](https://support.10xgenomics.com/single-cell-gene-expression/datasets/6.0.0/SC3_v3_NextGem_DI_Neurons_5K_SC3_v3_NextGem_DI_Neurons_5K)

3. Find genes associated to DA peaks and run GO over-enrichment / gene set enrichment analysis  

4. Compare `Signac` processing and `SingleCellExperiment` processing

## 1. Process mouse E18 brain 5K scATACseq data

### Download data 

scATACseq for mouse E18 5K brain can be downloaded from 
[10X Genomics](https://support.10xgenomics.com/single-cell-atac/datasets/1.2.0/atac_v1_E18_brain_fresh_5k). 

```{sh}
mkdir -p data/Homework_2/scATAC
curl https://cf.10xgenomics.com/samples/cell-atac/1.2.0/atac_v1_E18_brain_fresh_5k/atac_v1_E18_brain_fresh_5k_analysis.tar.gz -o data/Homework_2/scATAC/atac_v1_E18_brain_fresh_5k_analysis.tar.gz
curl https://cf.10xgenomics.com/samples/cell-atac/1.2.0/atac_v1_E18_brain_fresh_5k/atac_v1_E18_brain_fresh_5k_filtered_peak_bc_matrix.h5 -o data/Homework_2/scATAC/atac_v1_E18_brain_fresh_5k_filtered_peak_bc_matrix.h5
curl https://cf.10xgenomics.com/samples/cell-atac/1.2.0/atac_v1_E18_brain_fresh_5k/atac_v1_E18_brain_fresh_5k_filtered_tf_bc_matrix.h5 -o data/Homework_2/scATAC/atac_v1_E18_brain_fresh_5k_filtered_tf_bc_matrix.h5
curl https://cf.10xgenomics.com/samples/cell-atac/1.2.0/atac_v1_E18_brain_fresh_5k/atac_v1_E18_brain_fresh_5k_fragments.tsv.gz -o data/Homework_2/scATAC/atac_v1_E18_brain_fresh_5k_fragments.tsv.gz
curl https://cf.10xgenomics.com/samples/cell-atac/1.2.0/atac_v1_E18_brain_fresh_5k/atac_v1_E18_brain_fresh_5k_fragments.tsv.gz.tbi -o data/Homework_2/scATAC/atac_v1_E18_brain_fresh_5k_fragments.tsv.gz.tbi
curl https://cf.10xgenomics.com/samples/cell-atac/1.2.0/atac_v1_E18_brain_fresh_5k/atac_v1_E18_brain_fresh_5k_peak_annotation.tsv -o data/Homework_2/scATAC/atac_v1_E18_brain_fresh_5k_peak_annotation.tsv
curl https://cf.10xgenomics.com/samples/cell-atac/1.2.0/atac_v1_E18_brain_fresh_5k/atac_v1_E18_brain_fresh_5k_peaks.bed -o data/Homework_2/scATAC/atac_v1_E18_brain_fresh_5k_peaks.bed
curl https://cg.10xgenomics.com/samples/cell-atac/1.2.0/atac_v1_E18_brain_fresh_5k/atac_v1_E18_brain_fresh_5k_possorted_bam.bam -o data/Homework_2/scATAC/atac_v1_E18_brain_fresh_5k_possorted_bam.bam
curl https://cf.10xgenomics.com/samples/cell-atac/1.2.0/atac_v1_E18_brain_fresh_5k/atac_v1_E18_brain_fresh_5k_possorted_bam.bam.bai -o data/Homework_2/scATAC/atac_v1_E18_brain_fresh_5k_possorted_bam.bam.bai
curl https://cf.10xgenomics.com/samples/cell-atac/1.2.0/atac_v1_E18_brain_fresh_5k/atac_v1_E18_brain_fresh_5k_singlecell.csv -o data/Homework_2/scATAC/atac_v1_E18_brain_fresh_5k_singlecell.csv
curl https://cf.10xgenomics.com/samples/cell-atac/1.2.0/atac_v1_E18_brain_fresh_5k/atac_v1_E18_brain_fresh_5k_web_summary.html -o data/Homework_2/scATAC/atac_v1_E18_brain_fresh_5k_web_summary.html
curl https://cf.10xgenomics.com/samples/cell-atac/1.2.0/atac_v1_E18_brain_fresh_5k/atac_v1_E18_brain_fresh_5k_cloupe.cloupe -o data/Homework_2/scATAC/atac_v1_E18_brain_fresh_5k_cloupe.cloupe
```

### Import data in R 

First, the count matrix has to be imported. Use `Seurat` to do import the filtered *peak* count matrix. 
How many peaks were found? 

```{r}
library(Signac)
library(Seurat)
library(tidyverse)
library(BiocParallel)
library(plyranges)

## -- Read brain ATAC-seq counts
brain_assay <- Signac::CreateChromatinAssay(
    counts = Seurat::Read10X_h5("data/Homework_2/scATAC/atac_v1_E18_brain_fresh_5k_filtered_peak_bc_matrix.h5"),
    fragments = 'data/Homework_2/scATAC/atac_v1_E18_brain_fresh_5k_fragments.tsv.gz',
    sep = c(":", "-"),
    genome = "mm10",
    min.cells = 1
)
brain_assay
dim(brain_assay)
```

Cell metadata also needs to be read. Metadata is stored in a regular `csv` file. 
You can import it in R using `read.csv()` from `base R`, but `vroom` package wil take 
care of finding the best and fastest way to read it. It is especially useful for reading 
large rectangular data in R. 

How many cell barcode are there in total, in the `*_singlecell.csv` file? Should 
we take this into consideration and filter some out? 

```{r}
## -- Read metadata
metadata <- vroom::vroom(
    file = "data/Homework_2/scATAC/atac_v1_E18_brain_fresh_5k_singlecell.csv",
    col_names = TRUE
) %>% 
    filter(barcode != 'NO_BARCODE') %>% 
    mutate(cell = barcode) %>%
    filter(cell %in% colnames(brain_assay)) %>%
    column_to_rownames('barcode') %>% 
    as.data.frame() 

```

Now that cells and counts over peaks / cell are created, one can bind them together 
into a `Seurat` object. Try and create a standard `Seurat` object containing 
the `chromatinAssay` object and corresponding cell metadata. 

```{r}
## -- Create Seurat object
brain <- Seurat::CreateSeuratObject(
    counts = brain_assay,
    assay = 'peaks',
    project = 'ATAC',
    meta.data = metadata
)

## -- Add genome annotations
annotations <- GetGRangesFromEnsDb(ensdb = EnsDb.Mmusculus.v79::EnsDb.Mmusculus.v79)
seqlevelsStyle(annotations) <- 'UCSC'
genome(annotations) <- "mm10"
Annotation(brain) <- annotations
```

### QC data 

> Nucleosome signal

Few QC controls can be done. The `nucleosome signal` metric estimates the 
ratio of nucleosome-spanning fragments vs mononucleosomal fragments, for each cell. 
A low ratio (e.g. `< 4`) indicates enrichment of fragments over an open 
chromatin locus. 

```{r}
## -- Nucleosome signal 
brain <- Signac::NucleosomeSignal(object = brain)
brain$nucleosome_group <- ifelse(brain$nucleosome_signal > 4, 'NS > 4', 'NS < 4')
range(brain$nucleosome_signal)
table(brain$nucleosome_group)
```

Check the number of fragments / cell overlapping a known DHS locus (check `brain` 
pre-computed metrics to find the corresponding metric), for cells with varying 
`nucleosome_signal` scores.

```{r}
p<- tibble(dnase = brain$DNase_sensitive_region_fragments, nuc_signal = brain$nucleosome_signal) %>% 
    mutate(
        nuc_group = round(nuc_signal / 5)*5 + 1, 
        nuc_group = factor(nuc_group, seq(1, max(nuc_group)))
    ) %>% 
    ggplot(aes(x = nuc_group, y = dnase)) + 
    geom_boxplot() + 
    labs(x = 'Av. nucleosome signal')
```

> Fragment size distribution

We have covered this already in previous exercises focusing on bulk ATAC. 
It is a metric that one can also evaluate in scATACseq. 
For this, one can import the fragments (from `*_fragments.tsv.gz` file) in R. 
Here again, we highly recommend the use of `vroom` package, as the fragment file 
can be several Gb big!

Try plotting the fragment width for the first 10M fragments. 

```{r}
## -- Frag. size distribution
GetFragmentData(Fragments(brain)[[1]])
frags <- vroom::vroom(
        GetFragmentData(Fragments(brain)[[1]]), 
        col_names = FALSE, 
        n_max = 10000000
    ) %>% 
    setNames(c('seqnames', 'start', 'end', 'cell', 'cluster'))
p <-frags %>% 
    mutate(width = end - start) %>%
    ggplot(aes(x = width)) + 
    geom_bar() + 
    xlim(c(0, 800))
```

> TSSES 

This metric has also already been previously mentionned. It essentially represents 
the signal-to-noise ratio at TSSs. 
It can be estimated from a `Seurat` object using the `Signac::TSSEnrichment()` function. 

```{r}
## -- TSS enrichment
brain <- TSSEnrichment(brain, fast = TRUE)
quantile(brain$TSS.enrichment, probs = seq(0, 1, 0.1))
```

We can also plot the average coverage at TSSs. Let's plot it for the first 
10M fragments

```{r}
p<- as_granges(frags) %>% 
    resize(1, fix = 'start') %>% 
    coverage() %>%
    '['(target) %>% 
    data.matrix() %>%
    as_tibble() %>% 
    setNames(seq(-999.5, 999.5, length.out = 2000)) %>%
    mutate(peak = seq(1, nrow(.)), strand = as.character(strand(target))) %>% 
    pivot_longer(-c(peak, strand), names_to = 'distance', values_to = 'cnt') %>% 
    mutate(distance = as.numeric(distance)) %>% 
    mutate(distance = ifelse(strand == '-', -distance, distance)) %>%
    select(-peak, -strand) %>%
    group_by(distance) %>%
    tally(cnt) %>% 
    ggplot(aes(x = distance, y = n)) + 
    geom_line() +
    geom_smooth(method = 'loess', span = 0.05) + 
    theme_minimal() + 
    theme(legend.position = 'none') + 
    labs(title = 'Fragment "cut" sites', x = 'Distance from TSS', y = '# of cut sites')
```

### Subsetting data

Check the distribution of values for important metrics related to scATAC 
quality. Notably, the # and % of fragments overlapping identified peaks or 
blacklist regions. 

```{r}
quantile(brain$peak_region_fragments, probs = seq(0, 1, 0.1))
quantile(brain$blacklist_region_fragments, probs = seq(0, 1, 0.1))
brain$FRiP <- brain$peak_region_fragments / brain$passed_filters * 100
quantile(brain$FRiP, probs = seq(0, 1, 0.1))
brain$FRiBl <- brain$blacklist_region_fragments / brain$peak_region_fragments
quantile(brain$FRiBl, probs = seq(0, 1, 0.1))
```

Set sensible thresholds to remove cells which do not seem to be high-quality, 
then subset the `brain` Seurat object using these thresholds.

```{r}
brain <- subset(
    brain,
    peak_region_fragments > 2000 & # Keep cells with high number of fragments in peaks
    peak_region_fragments < 100000 & # Remove cells with number of fragments in peaks too high
    FRiP > 50 & # Keep cells with high FRiP 
    FRiBl < 0.05 & # Keep cells with low FRiBl
    nucleosome_signal < 4 & # Remove cells with high nucleosome signal
    TSS.enrichment > 2 # Keep cells with good TSSES
)
```

Don't forget to also subset some peaks (which may have been identified from the 
low-quality cells). 

```{r}
brain <- brain[
    rowSums(GetAssayData(brain, slot = "counts") > 0) > 10, # Remove peaks detected in less than 10 cells
]
```

### Normalizing and reduce dimensionality data 

Normalization is done using a TF-IDF approach, rather than traditional log-normalization, 
due to excessive sparsity of the scATAC data. 

```{r}
GetAssayData(brain, 'counts')[1:10, 1:10]
GetAssayData(brain, 'data')[1:10, 1:10]
brain <- Signac::RunTFIDF(brain)
GetAssayData(brain, 'counts')[1:10, 1:10]
GetAssayData(brain, 'data')[1:10, 1:10]
```

Dimensionality reduction can then be performed using the normalized data. 

```{r}
brain <- Signac::FindTopFeatures(brain, min.cutoff = 'q50') # Find variable features
brain <- Signac::RunSVD(object = brain) 
```

### Clustering 

Now that cells are embedded in a lower dimensional space, they can be graph-clustered just like "regular" cells from scRNAseq. 

```{r}
brain <- FindNeighbors(object = brain, reduction = 'lsi', dims = 2:30)
brain <- FindClusters(object = brain, algorithm = 3, resolution = 1.2, verbose = FALSE)
```

### Visualization 

```{r}
brain <- RunTSNE(brain, reduction = 'lsi', dims = 2:30)
brain <- RunUMAP(brain, reduction = 'lsi', dims = 2:30)
p1 <- Seurat::DimPlot(brain, reduction = 'tsne')
p2 <- CoveragePlot(
    object = brain,
    region = "Tjp1", # Or Gad2 (interneurons), Cq1a (Microglia), Gfap (astrocytes), Tubb3 (neurons)
    annotation = TRUE,
    peaks = FALSE
)
```

## 2. Integrate scATACseq and scRNAseq 

### Download data 

scRNAseq for mouse E18 5K brain can be downloaded from 
[10X Genomics](https://support.10xgenomics.com/single-cell-gene-expression/datasets/6.0.0/SC3_v3_NextGem_DI_Neurons_5K_SC3_v3_NextGem_DI_Neurons_5K). 

```{sh}
mkdir data/Homework_2/scRNA
curl https://cf.10xgenomics.com/samples/cell-exp/6.0.0/SC3_v3_NextGem_DI_Neurons_5K_SC3_v3_NextGem_DI_Neurons_5K/SC3_v3_NextGem_DI_Neurons_5K_SC3_v3_NextGem_DI_Neurons_5K_web_summary.html -o data/Homework_2/scRNA/SC3_v3_Neurons_5K_web_summary.html
curl https://cf.10xgenomics.com/samples/cell-exp/6.0.0/SC3_v3_NextGem_DI_Neurons_5K_SC3_v3_NextGem_DI_Neurons_5K/SC3_v3_NextGem_DI_Neurons_5K_SC3_v3_NextGem_DI_Neurons_5K_metrics_summary.csv -o data/Homework_2/scRNA/SC3_v3_Neurons_5K_metrics_summary.csv
curl https://cf.10xgenomics.com/samples/cell-exp/6.0.0/SC3_v3_NextGem_DI_Neurons_5K_SC3_v3_NextGem_DI_Neurons_5K/SC3_v3_NextGem_DI_Neurons_5K_SC3_v3_NextGem_DI_Neurons_5K_count_cloupe.cloupe -o data/Homework_2/scRNA/SC3_v3_Neurons_5K_count_cloupe.cloupe
curl https://cf.10xgenomics.com/samples/cell-exp/6.0.0/SC3_v3_NextGem_DI_Neurons_5K_SC3_v3_NextGem_DI_Neurons_5K/SC3_v3_NextGem_DI_Neurons_5K_SC3_v3_NextGem_DI_Neurons_5K_count_sample_alignments.bam -o data/Homework_2/scRNA/SC3_v3_Neurons_5K_count_sample_alignments.bam
curl https://cf.10xgenomics.com/samples/cell-exp/6.0.0/SC3_v3_NextGem_DI_Neurons_5K_SC3_v3_NextGem_DI_Neurons_5K/SC3_v3_NextGem_DI_Neurons_5K_SC3_v3_NextGem_DI_Neurons_5K_count_sample_alignments.bam.bai -o data/Homework_2/scRNA/SC3_v3_Neurons_5K_count_sample_alignments.bam.bai
curl https://cf.10xgenomics.com/samples/cell-exp/6.0.0/SC3_v3_NextGem_DI_Neurons_5K_SC3_v3_NextGem_DI_Neurons_5K/SC3_v3_NextGem_DI_Neurons_5K_SC3_v3_NextGem_DI_Neurons_5K_count_sample_barcodes.csv -o data/Homework_2/scRNA/SC3_v3_Neurons_5K_count_sample_barcodes.csv
curl https://cf.10xgenomics.com/samples/cell-exp/6.0.0/SC3_v3_NextGem_DI_Neurons_5K_SC3_v3_NextGem_DI_Neurons_5K/SC3_v3_NextGem_DI_Neurons_5K_SC3_v3_NextGem_DI_Neurons_5K_count_sample_feature_bc_matrix.h5 -o data/Homework_2/scRNA/SC3_v3_Neurons_5K_count_sample_feature_bc_matrix.h5
curl https://cf.10xgenomics.com/samples/cell-exp/6.0.0/SC3_v3_NextGem_DI_Neurons_5K_SC3_v3_NextGem_DI_Neurons_5K/SC3_v3_NextGem_DI_Neurons_5K_SC3_v3_NextGem_DI_Neurons_5K_count_sample_feature_bc_matrix.tar.gz -o data/Homework_2/scRNA/SC3_v3_Neurons_5K_count_sample_feature_bc_matrix.tar.gz
curl https://cf.10xgenomics.com/samples/cell-exp/6.0.0/SC3_v3_NextGem_DI_Neurons_5K_SC3_v3_NextGem_DI_Neurons_5K/SC3_v3_NextGem_DI_Neurons_5K_SC3_v3_NextGem_DI_Neurons_5K_count_sample_molecule_info.h5 -o data/Homework_2/scRNA/SC3_v3_Neurons_5K_count_sample_molecule_info.h5
curl https://cf.10xgenomics.com/samples/cell-exp/6.0.0/SC3_v3_NextGem_DI_Neurons_5K_SC3_v3_NextGem_DI_Neurons_5K/SC3_v3_NextGem_DI_Neurons_5K_SC3_v3_NextGem_DI_Neurons_5K_count_analysis.tar.gz -o data/Homework_2/scRNA/SC3_v3_Neurons_5K_count_analysis.tar.gz
```

### Process data 

```{r}
library(tidyverse)
library(Seurat)
brain_rna <- Read10X_h5("data/Homework_2/scRNA/SC3_v3_Neurons_5K_count_sample_feature_bc_matrix.h5") %>% 
    CreateSeuratObject() %>% 
    SCTransform() %>% 
    RunPCA() %>% 
    FindNeighbors() %>% 
    FindClusters() %>% 
    RunTSNE()
brain_rna <- RunUMAP(brain_rna, reduction = 'pca', dims = 1:50)
p <- DimPlot(brain_rna, reduction = "umap")
```

### Annotate scRNAseq data with public databases

```{r}
ref <- scRNAseq::ZeiselBrainData() %>% scuttle::logNormCounts()
ref$level1class[ref$level1class %in% c('pyramidal CA1', 'pyramidal SS')] <- 'pyramidal'
pred <- SingleR::SingleR(
    test = GetAssayData(brain_rna, 'data'), 
    ref = ref, 
    labels = ref$level1class, 
    BPPARAM = BiocParallel::MulticoreParam(workers = 12)
)
brain_rna$annotation <- pred$pruned.labels
p <- DimPlot(brain_rna, reduction = "umap", group.by = 'annotation')
```

### Transfer annotations from scRNAseq to scATACseq datasets

```{r}
## - Compute an "activity score" for each gene in each cell (based on ATAC counts) and store it as a new assay
gene_activities <- GeneActivity(brain, features = VariableFeatures(brain_rna))
brain@assays["gene_activity"] <- CreateAssayObject(counts = gene_activities)

## - Normalize this new assay with SCTransform
GetAssayData(brain, 'counts')[1:10, 1:10]
brain <- SCTransform(brain, assay = 'gene_activity')

## - Normalize this new assay with SCTransform
anchors <- FindTransferAnchors(
    reference = brain_rna, 
    query = brain, 
    features = VariableFeatures(object = brain_rna),
    reference.assay = "SCT", 
    query.assay = "SCT", 
    reduction = "cca"
)

## - Transfer annotations from scRNA to scATAC
pred_labels <- TransferData(
    anchorset = anchors, 
    refdata = brain_rna$annotation,
    weight.reduction = brain[['lsi']], 
    dims = 2:30
)

## - Save transferred annotations
brain$transferred_annotation <- pred_labels$predicted.id
brain$transferred_annotation[pred_labels$prediction.score.max < 0.8] <- NA
p <- Seurat::DimPlot(brain, reduction = 'tsne', group.by = 'transferred_annotation')
```

## 3. Functional analysis of DA peaks

### Extract peaks enriched in astrocytes vs. others 

```{r}
## - Find astrocyte-specific peaks 
Idents(brain) <- brain$transferred_annotation
astrocyte_peaks <- Seurat::FindMarkers(
    object = brain,
    assay = 'peaks', 
    ident.1 = 'astrocytes_ependymal',
    ident.2 = unique(Idents(brain)[Idents(brain) != 'astrocytes_ependymal']),
    min.pct = 0.15, 
    only.pos = TRUE
)
```

### Find nearest gene for each astrocyte-specific peak

```{r}
## - Link astrocyte peaks to nearest genes
mm_genes <- ensembldb::genes(EnsDb.Mmusculus.v79::EnsDb.Mmusculus.v79) %>% filter(gene_biotype == 'protein_coding')
seqlevelsStyle(mm_genes) <- 'UCSC'
astrocyte_genes <- astrocyte_peaks %>% 
    filter(p_val_adj <= 0.05, avg_log2FC > 1, pct.1 > 0.2) %>% 
    rownames() %>% 
    str_replace('-', ':') %>% 
    GRanges() %>% 
    join_nearest(mm_genes) %>% 
    as_tibble() %>% pull(symbol) %>% 
    unique()
```

### Perform GO over-representation analysis

```{r}
## - Run GO analysis
goa <- gprofiler2::gost(astrocyte_genes, organism = 'mmusculus')
goa$result %>% as_tibble() %>% filter(p_value <= 0.05, grepl('GO:', source)) %>% arrange(p_value) %>% print(n = 40)
```

### Repeat for microgli-annotated cluster

```{r}
## - Find microglia-specific peaks 
microglia_peaks <- Seurat::FindMarkers(
    object = brain,
    assay = 'peaks', 
    ident.1 = 'microglia',
    ident.2 = unique(Idents(brain)[Idents(brain) != 'microglia']),
    min.pct = 0.15, 
    only.pos = TRUE
)

## - Link microglia peaks to nearest genes
microglia_genes <- microglia_peaks %>% 
    arrange(desc(avg_log2FC)) %>% 
    filter(p_val_adj <= 0.05, avg_log2FC > 1, pct.1 > 0.2) %>% 
    rownames() %>% 
    str_replace('-', ':') %>% 
    GRanges() %>% 
    join_nearest(mm_genes) %>% 
    as_tibble() %>% pull(symbol) %>% 
    unique()

all_genes <- rownames(brain) %>% 
    str_replace('-', ':') %>% 
    GRanges() %>% 
    join_nearest(mm_genes) %>% 
    as_tibble() %>% pull(symbol) %>% 
    unique()

## - Run GO analysis
goa <- gprofiler2::gost(microglia_genes, organism = 'mmusculus')
goa$result %>% as_tibble() %>% filter(p_value <= 0.05, grepl('GO:', source)) %>% arrange(p_value) %>% print(n = 40)
```

```{r}
cov_plot <- CoveragePlot(
  object = pbmc,
  region = "chr2-87011729-87035519",
  annotation = FALSE,
  peaks = FALSE
)
cov_plot
```

## 4. [Optional] Compare Signac and SingleCellExperiment workflows

You will need several packages which may not be already installed. 

- DropletUtils

```{r}
# Import data
brain_sce <- DropletUtils::read10xCounts('data/Homework_2/scATAC/atac_v1_E18_brain_fresh_5k_filtered_peak_bc_matrix.h5')
# QC data 

# Subset data

# Normalize data 

# Reduce dimensionality

# Clustering 

# Visualization 

```