---
title: "Homework 3"
output:
  blogdown::html_page:
    highlight: tango
---

```{r echo = FALSE}
knitr::opts_chunk$set(eval = FALSE)
```

## Goals

1. Import, filter, normalize and plot cells from human hematopoiesis scATACseq dataset

2. Compute gene activity scores and check known markers  

4. Run chromVAR to identify TF activity in each cell cluster  

## 1. Process human hematopoiesis dataset

### Download data 

The data comes from Satpathy et al., Nat. Biotechnol. 2019 ([DOI: 10.1038/s41587-019-0206-z](https://doi.org/10.1038/s41587-019-0206-z)). 
Counts data is available from GEO (GSE: `GSE129785`). 

Download the files related to scATACseq of all human hematopoiesis cells. 

```{sh}
mkdir -p data/Homework_3/scATAC/
curl https://ftp.ncbi.nlm.nih.gov/geo/series/GSE129nnn/GSE129785/suppl/GSE129785_scATAC-Hematopoiesis-All.cell_barcodes.txt.gz -o data/Homework_3/scATAC/barcodes.tsv.gz
curl https://ftp.ncbi.nlm.nih.gov/geo/series/GSE129nnn/GSE129785/suppl/GSE129785_scATAC-Hematopoiesis-All.mtx.gz -o data/Homework_3/scATAC/matrix.mtx.gz
curl https://ftp.ncbi.nlm.nih.gov/geo/series/GSE129nnn/GSE129785/suppl/GSE129785_scATAC-Hematopoiesis-All.peaks.txt.gz -o data/Homework_3/scATAC/features.tsv.gz
```

### Process data wih Signac

#### Import data 

Notice how the count matrix is in a different format than previously seen. 
You will frequently encounter this type of situation. Analysis packages often provide 
import functions for the different input formats. Check `Seurat` documentation if needed, 
to identify which function can be used to import a `mtx` format. 

```{r}
cnts <- Seurat::Read10X('data/Homework_3/scATAC/')
```

Does it work? Why? The most likely reason is that the count matrix is not 
structured exactly like the one which comes straight out of a `cellranger count` 
pipeline. It is a very frequent issue occurring when one attempts to analyse 
public data. 

The alternative, longer route is to import each file separately, 
make sure they are compatible with each other, and merge them 
manually into a `SeuratChromatinAssay`. 

```{r}
## - Load required libraries 
library(tidyverse)
library(plyranges)
library(Seurat)
library(Signac)

## - Import counts, barcodes and features separately
barcode <- file.path('data/Homework_3/scATAC/', "barcodes.tsv.gz")
features <- file.path('data/Homework_3/scATAC/', "features.tsv.gz")
matrix <- file.path('data/Homework_3/scATAC/', "matrix.mtx.gz")
cnts <- Matrix::readMM('data/Homework_3/scATAC/matrix.mtx.gz')
barcodes <- vroom::vroom('data/Homework_3/scATAC/barcodes.tsv.gz') %>% as.data.frame()
features <- readLines('data/Homework_3/scATAC/features.tsv.gz')[-1] %>% 
    str_replace('_', ':') %>% 
    str_replace('_', '-') %>% 
    GRanges() %>% 
    mutate(peak = as.character(.) %>% str_replace(':', '-'))

## - Make sure that the row/colnames for the count matrix match those of 
#       barcodes and features
colnames(cnts) <- rownames(barcodes) <- barcodes$Group_Barcode
rownames(cnts) <- names(features) <- features$peak

## - Get human gene annotations (hg19/GRCh37) to feed it into the future `SeuratChromatinAssay`
annotations <- GetGRangesFromEnsDb(ensdb = EnsDb.Hsapiens.v75::EnsDb.Hsapiens.v75)
seqlevelsStyle(annotations) <- 'UCSC'
genome(annotations) <- "hg19"

## - Create Chromatin Assay
assay <- Signac::CreateChromatinAssay(
    counts = cnts,
    ranges = features,
    genome = "hg19",
    min.cells = 10, 
    min.features = 10, 
    annotation = annotations
)
hemato <- Seurat::CreateSeuratObject(
    counts = assay,
    assay = 'ATAC',
    meta.data = barcodes
)
```

You will find that `CreateChromatinAssay()` function takes a long time to run. 
This is because it internally computes metrics from the count matrix. You can 
check the cell metadata to see which ones were calculated (on top of the ones 
which come from the `barcodes` data frame). 

```{r}
head(hemato[[]])
colnames(hemato[[]])[!colnames(hemato[[]]) %in% colnames(barcodes)]
```

#### Process data 

Is the `fragments` file (which should exist for any scATACseq experiment) available 
for this experiment? Check the related GEO webpage. 

For which analysis are the `fragments` essential, exactly? Can we still 
perform normalization/clustering/annotation without them? And motif enrichment analysis? 

Since we don't have the `fragments` file at hand, most of the QC steps are not 
available (e.g. `TSSEnrichment`, `NucleosomeSignal` or fragment size distribution). 

(Note that the `fragments` file is actually available as raw data from GEO. 
Feel free to download it and perform further QC).

Still, we do have other pre-computed metrics (e.g. `FRIP` and `depth`). 
Check these metrics and filter the `Seurat` object (cells and features) 
as deemed appropriate. 

```{r}
## - Filter data 
quantile(hemato$depth, seq(0, 1, 0.1))
quantile(hemato$nFeature_ATAC, seq(0, 1, 0.1))
quantile(hemato$nCount_ATAC, seq(0, 1, 0.1))
quantile(hemato$FRIP, seq(0, 1, 0.1))
hemato <- subset(hemato, subset = depth > 5000 & depth < 200000) 
hemato <- subset(hemato, subset = nCount_ATAC > 3000 & nCount_ATAC < 100000) 
hemato <- subset(hemato, subset = nFeature_ATAC > 1000 & nFeature_ATAC < 20000) 
hemato <- subset(hemato, subset = FRIP > 0.30 & FRIP < 20000) 

## - Remove peaks with low coverage
hemato <- hemato[rowSums(GetAssayData(hemato, slot = "counts")) > 10, ]
hemato <- hemato[rowSums(GetAssayData(hemato, slot = "counts") > 0) > 10, ]
```

Now that the dataset is filtered, one can normalize (by using TF-IDF 
approach) then further reduce the dimensionality for visualization purposes. 

```{r}
## - Normalize data 
hemato <- Signac::RunTFIDF(hemato) 

## - Reduce dimensionality
hemato <- Signac::FindTopFeatures(hemato, min.cutoff = 'q50') 
hemato <- Signac::RunSVD(hemato) 

## - Label clusters according to the original publication
clusters <- c(
    'Cluster1' = 'HSC',
    'Cluster2' = 'MEP',
    'Cluster3' = 'CMP',
    'Cluster4' = 'LMPP',
    'Cluster5' = 'CLP',
    'Cluster6' = 'Pro-B',
    'Cluster7' = 'Pre-B',
    'Cluster8' = 'GMP',
    'Cluster9' = 'MDP',
    'Cluster10' = 'pDC',
    'Cluster11' = 'cDC',
    'Cluster12' = 'Mono.',
    'Cluster13' = 'Mono.',
    'Cluster14' = 'Nai. B',
    'Cluster15' = 'Mem. B',
    'Cluster16' = 'Plasma',
    'Cluster17' = 'Baso.',
    'Cluster18' = 'Imm. NK',
    'Cluster19' = 'Mat. NK',
    'Cluster20' = 'Mat. NK',
    'Cluster21' = 'Nai. CD4',
    'Cluster22' = 'Nai. CD4',
    'Cluster23' = 'Nai. Treg',
    'Cluster24' = 'Mem. CD4',
    'Cluster25' = 'Treg',
    'Cluster26' = 'Nai. CD8',
    'Cluster27' = 'Nai. CD8',
    'Cluster28' = 'Nai. CD8',
    'Cluster29' = 'Mem. CD8',
    'Cluster30' = 'Mem. CD8',
    'Cluster31' = 'Gamm. Del. T'
)
hemato$renamed_clusters <- clusters[hemato$Clusters]

## - Visualize data 
hemato <- Seurat::RunUMAP(hemato, reduction = 'lsi', dims = 2:30)
p <- DimPlot(hemato, group.by = 'renamed_clusters') + coord_fixed(ratio = 1)
```

What can you observe in the `UMAP` projection of the dataset? Comment on the 
separation of some cell types into different spatially-resolved clusters. 

## 2. Compute gene activity scores

Here again, the absence of `fragments` complicates the computation of gene activity scores. 
Indeed, Signac's `GeneActivity()` function does require scATACseq fragment information. 

Without actual fragments, can you think of a way to still compute a 'gene activity'-like score, 
approximating ~ the sum of scATACseq fragments overlapping a gene? 

One way to do this is to: 

1. Extend gene annotations to cover 2000bp upstream of TSS
2. For each gene, link it to all the peaks overlapping it to create a gene <-> peak relationship table
3. Create a joint data table containing [ gene | peak | cell | count ]
4. Group the data table by `cell + gene` and sum up the peak counts for each pair
5. Transform this count table into a SparseMatrix
6. Create the new `Seurat` assay containing "gene activity"-like scores and normalize it

- Step 1: 

```{r}
## - Extend all the genes in hg19 so that it covers 1000 bp upstream of the genebody 
hg_genes_extended <- ensembldb::genes(EnsDb.Hsapiens.v75::EnsDb.Hsapiens.v75) %>% 
    filter(gene_biotype == 'protein_coding') %>% 
    keepStandardChromosomes(pruning.mode = 'coarse') %>% 
    filter(seqnames != 'MT') %>% 
    resize(width = width(.) + 2000, fix = 'start') %>% 
    shift(1000 * ifelse(strand(.) == '+', -1, 1)) %>%
    mutate(id = 1:length(.)) %>% 
    mutate(gene_name = make.unique(gene_name))
seqlevelsStyle(hg_genes_extended) <- 'UCSC'
```

- Step 2: 

```{r}
## - Find overlaps between all the scATAC peaks and the extended genes
ov <- findOverlaps(query = granges(hemato), subject = hg_genes_extended)

## - For each peak-gene pair, add peak name and gene name
ov <- as_tibble(ov) %>% 
    dplyr::rename(peaknb = queryHits, genenb = subjectHits) %>% 
    mutate(gene_name = hg_genes_extended$gene_name[genenb]) %>% 
    mutate(row = granges(hemato)$peak[peaknb])
```

- Step 3: 

This step requires to transform the original `counts` SparseMatrix into a tidy 
`data.table`, to enable fast `left_join` between (1) peak counts / cell and 
(2) gene <-> peak relationship table (`ov`)

```{r}
## - Extract counts for the gene-associated peaks
cnts <- GetAssayData(hemato, 'counts')[unique(ov$peaknb), ]

## - Transform the cnts SparseMatrix into a tidy dense long data.frame
tidy_cnts <- broom::tidy(cnts) 

## - Transform the dense long data.frame into a data.table for faster computation
tidy_cnts <- dtplyr::lazy_dt(tidy_cnts)

## - For each peak/cell count value, associate the corresponding gene
joint_data <- left_join(tidy_cnts, ov, by = 'row')
class(joint_data)
## WATCH OUT: You cannot see what is stored into `joint_data` as it 
## is a deferred data.table, i.e. it will only execute once you pipe it to 
## `as_tibble()`. Once you pipe it, the computation starts and takes a lot of time.
```

- Step 4: 

```{r}
## - Group the joint_table by gene + cell and sum up all the counts
gene_averaged_cnts <- group_by(joint_data, column, gene_name) %>% 
    summarize(n = sum(value)) %>% 
    as_tibble()
```

- Step 5: 

```{r}
## - To transform the data.table into SparseMatrix, it needs to be numerical
gene_averaged_cnts_num <- data.matrix(gene_averaged_cnts)

## - Transform into sparse matrix 
gene_averaged_cnts_sparse <- Matrix::sparseMatrix(
    i = gene_averaged_cnts_num[, 'gene_name'],
    j = gene_averaged_cnts_num[, 'column'],
    x = gene_averaged_cnts_num[, 'n'],
)

## - Add back the actual cell and genes names
colnames(gene_averaged_cnts_sparse) <- sort(unique(gene_averaged_cnts$column))
rownames(gene_averaged_cnts_sparse) <- sort(unique(gene_averaged_cnts$gene_name))
```

- Step 6: 

```{r}
## - Create new assay in hemato
hemato[['GENE']] <- CreateAssayObject(counts = gene_averaged_cnts_sparse)

## - Normalize this new assay with SCTransform
hemato <- SCTransform(hemato, assay = 'GENE')

## - Save the resulting object!
saveRDS(hemato, 'data/Homework_3/scATAC/hemato.rds')
```

Now the `Seurat` object contains 3 assays, the first one corresponding to the original scATACseq counts, 
and the second one containing gene activity pseudocounts, and the third one containing `SCT`-normalized 
data. One can now perform "gene differential expression"-like analysis using the `SCT` assay!

```{r}
## - Reduce dimensionality and visualize
hemato <- RunPCA(hemato) %>% RunUMAP(reduction = 'pca', dims = 1:50)
p <- FeaturePlot(hemato, features = 'CD14', reduction = "umap") # Or NCR1 (NKp46), CD8A, CD14, MS4A1

## - Find gene markers for Basophils
DefaultAssay(hemato) <- "SCT"
Idents(hemato) <- hemato$renamed_clusters
basophils_genes <- Seurat::FindMarkers(
    object = hemato,
    assay = 'GENE', 
    ident.1 = 'Baso.',
    ident.2 = unique(Idents(hemato)[Idents(hemato) != 'Baso.']),
    min.pct = 0.15, 
    only.pos = TRUE
)
```

## 3. chromVAR analysis

### Downsample data to 10,000 random cells and a subset of variable features

```{r}
DefaultAssay(hemato) <- 'ATAC'
hemato_sub <- hemato[, sample(colnames(hemato), size = 10000, replace = FALSE)]
hemato_sub <- FindVariableFeatures(hemato_sub, nfeatures = 20000)
```

### Get a SummarizedExperiment of scATACseq counts over peaks

```{r}
library(SummarizedExperiment)
sumExp <- SummarizedExperiment(
    assays = list('counts' = GetAssayData(hemato, assay = 'ATAC', slot = 'counts')), 
    rowRanges = GetAssayData(hemato, assay = 'ATAC', slot = 'ranges'), 
    colData = hemato[[]]
)
```

### Add GC bias to peaks 

```{r}
sumExp <- chromVAR::addGCBias(sumExp, genome = BSgenome.Hsapiens.UCSC.hg19::BSgenome.Hsapiens.UCSC.hg19)
```

### Search for motifs with high deviation from background distribution 

```{r}
library(BiocParallel)
register(MulticoreParam(workers = 16, progressbar = TRUE))
motifs <- TFBSTools::getMatrixSet(
    JASPAR2020::JASPAR2020,
    list(species = 9606, all_versions = FALSE)
)
motif_ix <- motifmatchr::matchMotifs(motifs, sumExp, genome = BSgenome.Hsapiens.UCSC.hg19::BSgenome.Hsapiens.UCSC.hg19)
bg <- chromVAR::getBackgroundPeaks(object = sumExp)
dev <- chromVAR::computeDeviations(object = sumExp, annotations = motif_ix, background_peaks = bg)
```

### Check TF enrichment in different cell types 

The `dev` `SummarizedExperiment` object can be added as a fourth assay to the 
original `hemato` `Seurat` object. This way, one can rely on `Seurat`-based 
plotting functions to plot cells in their preferred dimensional space, and color 
them using TF deviation scores computed with `chromVAR`. 

```{r}

```

One can also embed cells in a 2D-space based on the TF deviation scores themselves! 

```{r}
variability <- chromVAR::computeVariability(dev)
p <- chromVAR::plotVariability(variability, use_plotly = FALSE) 
tsne_results <- chromVAR::deviationsTsne(dev, threshold = 1.5)
p <- cowplot::plot_grid(
    chromVAR::plotDeviationsTsne(dev, tsne_results, sample_column = "celltype", shiny = FALSE)[[1]], 
    chromVAR::plotDeviationsTsne(dev, tsne_results, annotation_name = "PAX5", shiny = FALSE)[[1]], 
    chromVAR::plotDeviationsTsne(dev, tsne_results, annotation_name = "TCF7", shiny = FALSE)[[1]], 
    chromVAR::plotDeviationsTsne(dev, tsne_results, annotation_name = "CEBPA", shiny = FALSE)[[1]]
)
```

### Exercise session 6

- Try different thresholds for tSNE embeddings. Comment. 

```{txt}
set.seed(2021)
dev <- readRDS('data/scATAC_LSC-LMPP-mono/dev.rds')
p <- cowplot::plot_grid(
    chromVAR::plotDeviationsTsne(dev, chromVAR::deviationsTsne(dev, threshold = 1, perplexity = 20), annotation_name = "FOS", shiny = FALSE)[[1]],
    chromVAR::plotDeviationsTsne(dev, chromVAR::deviationsTsne(dev, threshold = 1.25, perplexity = 20), annotation_name = "FOS", shiny = FALSE)[[1]],
    chromVAR::plotDeviationsTsne(dev, chromVAR::deviationsTsne(dev, threshold = 1.5, perplexity = 20), annotation_name = "FOS", shiny = FALSE)[[1]],
    chromVAR::plotDeviationsTsne(dev, chromVAR::deviationsTsne(dev, threshold = 1.75, perplexity = 20), annotation_name = "FOS", shiny = FALSE)[[1]]
)
```
